Static Analysis by Path Focusing

### 2011-05-18 report/report.tex
added some missing code for a more correct construction of rho (when condition branchments depend on values of boolean Phi nodes)

\title{Static Analysis by Path Focusing}
\author{Julien Henry}
\date{2011}
\shorttitle{Static Analysis by Path Focusing}
\institute{Grenoble-INP}
\abstract{
	Program verification aims at statically discover properties on programs,
	such as the values that can take the different variables during execution.
	Abstract Interpretation is a technique that allows to compute an
	approximation of the set of these values, since it is impossible to compute
	the real set in general.
	This report takes place in the many attempts to improve the precision of
	static Analysis by Abstract Interpretation. It proposes a technique that
	takes benefit of SMT-solving to obtain more precise results at reasonable
	cost.
}
\resume{  
	La vérification de programme consiste à découvrir statiquement des
	propriétés sur des programmes, comme l'ensemble des valeurs que peuvent
	prendre les variables durant l'exécution. L'Interprétation Abstraite est une
	technique permettant de calculer une approximation de cet ensemble, le
	véritable ensemble étant impossible à calculer en général.
	Ce report s'inscrit dans la lignée des travaux visant à améliorer la
	précision de l'analyse par interprétation abstraite, et propose une
	technique tirant parti du SMT-solving pour obtenir de meilleurs résultats
	à un coût raisonnable.
}
\keywords{Static Program Analysis, Verification, Abstract Interpretation,
Invariant Generation, Path Focusing, SMT-Solving}
\motscles{Analyse Statique de Programmes, Vérification, Interprétation
Abstraite, Génération d'invariants, Découverte de Chemins, SMT-Solving}
\tutor{David Monniaux - Matthieu Moy}
\notes{some notes}
\webaddress{Julien.Henry@imag.fr}
\reporttype{M2R internship's report}

@@

Static analysis aims at automatically computing properties on programs, such as
the possible values of their variables during execution. This allows to show for
instance that a program will not overflow, will not divide by zero, to compute
loop invariants\dots
The main applications of static analysis are compilation-time optimisation, and
the proof of satefy properties in critical systems, such as avionics.

@@

Satisfiability Modulo Theory (SMT) solving is a field in computer science that
aims to decide the satisfiability of a logic formula containing Boolean
predicates and elements of a theory, such as linear arithmetic relations between
integers.
In this report, we present a novel refinement of the Abstract Interpretation
technique, that uses SMT-solving to guide the fixpoint iterations, so that it
temporarily focuses on a selected path in the control flow graph (the graph
expressing the semantics of the program), in order to
obtain more precise results at reasonable cost.

@@

Static analysis aims to discover some properties about programs. In this
example, one would compute the set of possible values for the variable $x$
during execution.
The graph at the right side represents the program. The nodes of the graph are
called \emph{program points}. We can compute for instance the set $Y_1$ of
values for $x$ at point $p_1$. This set depends on the sets $Y_0$ and $Y_2$,
since there are two edges arriving at $p_1$: the first one comes from $p_0$, and
the second one from $p_2$.

@@

Indeed, if $\Phi^\#(x) = x$, then $\Phi \circ \gamma (x) \subseteq \gamma(x)
= \gamma \circ \Phi^\#(x)$. Abstract interpretation technique aims at computing
one of these $x$. It computes the stationary limit of an ascending sequence
$(x_i)_{i \geq 0}$ defined by the induction:

@@

Linear Relation Analysis \cite{CH78} (also noted LRA) is a direct application of
Abstract Interpretation. It aims at computing an upper approximation of the
reachable states of a program containing numerical variables. The set of
possible assignments for the numerical variables is abstracted by convex
polyhedron. This technique discovers invariant linear relations between
the numerical variables at each control point of a program.

@@

The computation aims to find a solution for this system of abstract semantic
inequalities:

@@

Typically, to analyze a program with loops, the approach is to use the
widening operator, for example at the head of the loops, to ensure convergence.
Since this induces imprecision, another technique called \emph{acceleration}
aims to compute the exact effect of the loop when possible \cite{Gon07,GH06}. For instance, for
the following program, an invariant could be directly computed for the loop:

@@

In some cases, a loop can have a non-regular behaviour, especially when it has
several paths. Some of these paths can stay impossible for a while, and become
possible after a certain number of iterations. However, the widening operator
extrapolates without taking care of these new possible paths. This can induce a
loss of precision, that Lookahead widening aims to limit.

@@

	\subsubsection{Reachability problem}
	\label{reachability}
	Path focusing technique aims to discover which paths to focus on for the
	computation of the fixpoint iterations. 
	Since the multigraph is not constructed explicitly, the different paths
	starting and finishing on node in $P_R$ are not known. 
	The technique proposes to express the focus path as the solution of a
	satisfiability modulo theory (SMT) problem:

@@

	We apply the \emph{Path Focusing} technique to our running example from
	Figures \ref{runningexample}, \ref{multigraphtransformGopan} and
	\ref{multigraphGopan}.
	The control points in $P_R$ are $p_0$, $p_1$ and $p_7$. 
	The $Ŕho$ formula associated to this program is depicted in
	Figure \ref{rhoformula}.
	One computes
	iteratively the associate abstract values $X_0$, $X_1$ and $X_7$.

@@

	The technique aims to compute a disjunctive invariant for this control
	point: 
	$$\displaystyle\bigvee_{1 \leq j \leq m} X_{p,j}$$
	where $X_{p,j}$ is a conjunction of linear inequalities, i.e a convex
	polyhedra.

@@

@@

Another example of information that was missing is the set of live variables.
Indeed, in LLVM, live variables computation is processed after the register
allocation. Then, the data structure we work on doesn't have live variables
informations yet. To solve this problem, we coded our own pass, that aims to
compute these live variables after the transformation in SSA form.

@@

	Validation of softwares has become an essential domain in industry,
	especially those dealing with embedded systems, where a bug in the program
	can endanger the safety of people, or induce huge costs to fix. These
	domains constitute a privileged application field for formal verification
	techniques.
	Abstract Interpretation is a common used technique to do static analysis of
	programs. It aims to compute iteratively an over-approximation of the set of
	possible states of the program, and allows to prove properties such as
	``In this program, $x$ is always between 1 and 10'', ``The program never
	divides by zero'', etc. The objective is to obtain a result as precise as
	possible, by limiting the loss of precision during the analysis, induced for
	instance by the widening operator, or control-flow merges.

>>>>

\title{Static Analysis by Path Focusing}
\author{Julien Henry}
\date{2011}
\shorttitle{Static Analysis by Path Focusing}
\institute{Grenoble-INP}
\abstract{
	Program verification is aimed at statically discover properties on programs,
	such as the values that can take the different variables during execution.
	Abstract Interpretation is a technique that allows to compute an
	approximation of the set of these values, since it is impossible to compute
	the real set in general.
	This report takes place in the many attempts to improve the precision of
	static Analysis by Abstract Interpretation. It proposes a technique that
	takes benefit of SMT-solving to obtain more precise results at reasonable
	cost.
}
\resume{  
	La vérification de programme consiste à découvrir statiquement des
	propriétés sur des programmes, comme l'ensemble des valeurs que peuvent
	prendre les variables durant l'exécution. L'Interprétation Abstraite est une
	technique permettant de calculer une approximation de cet ensemble, le
	véritable ensemble étant impossible à calculer en général.
	Ce report s'inscrit dans la lignée des travaux visant à améliorer la
	précision de l'analyse par interprétation abstraite, et propose une
	technique tirant parti du SMT-solving pour obtenir de meilleurs résultats
	à un coût raisonnable.
}
\keywords{Static Program Analysis, Verification, Abstract Interpretation,
Invariant Generation, Path Focusing, SMT-Solving}
\motscles{Analyse Statique de Programmes, Vérification, Interprétation
Abstraite, Génération d'invariants, Découverte de Chemins, SMT-Solving}
\tutor{David Monniaux - Matthieu Moy}
\notes{some notes}
\webaddress{Julien.Henry@imag.fr}
\reporttype{M2R internship's report}

@@

Static analysis is aimed at automatically computing properties on programs, such as
the possible values of their variables during execution. This allows to show for
instance that a program will not overflow, will not divide by zero, to compute
loop invariants\dots
The main applications of static analysis are compilation-time optimisation, and
the proof of satefy properties in critical systems, such as avionics.

@@

Satisfiability Modulo Theory (SMT) solving is a field in computer science that
is aimed to decide the satisfiability of a logic formula containing Boolean
predicates and elements of a theory, such as linear arithmetic relations between
integers.
In this report, we present a novel refinement of the Abstract Interpretation
technique, that uses SMT-solving to guide the fixpoint iterations, so that it
temporarily focuses on a selected path in the control flow graph (the graph
expressing the semantics of the program), in order to
obtain more precise results at reasonable cost.

@@

Static analysis is aimed at discovering some properties about programs. In this
example, one would compute the set of possible values for the variable $x$
during execution.
The graph at the right side represents the program. The nodes of the graph are
called \emph{program points}. We can compute for instance the set $Y_1$ of
values for $x$ at point $p_1$. This set depends on the sets $Y_0$ and $Y_2$,
since there are two edges arriving at $p_1$: the first one comes from $p_0$, and
the second one from $p_2$.

@@

Indeed, if $\Phi^\#(x) = x$, then $\Phi \circ \gamma (x) \subseteq \gamma(x)
= \gamma \circ \Phi^\#(x)$. Abstract interpretation technique is aimed at computing
one of these $x$. It computes the stationary limit of an ascending sequence
$(x_i)_{i \geq 0}$ defined by the induction:

@@

Linear Relation Analysis \cite{CH78} (also noted LRA) is a direct application of
Abstract Interpretation. It is aimed at computing an upper approximation of the
reachable states of a program containing numerical variables. The set of
possible assignments for the numerical variables is abstracted by convex
polyhedron. This technique discovers invariant linear relations between
the numerical variables at each control point of a program.

@@

The computation is aimed at finding a solution for this system of abstract semantic
inequalities:

@@

Typically, to analyze a program with loops, the approach is to use the
widening operator, for example at the head of the loops, to ensure convergence.
Since this induces imprecision, another technique called \emph{acceleration}
is aimed at computing the exact effect of the loop when possible \cite{Gon07,GH06}. For instance, for
the following program, an invariant could be directly computed for the loop:

@@

In some cases, a loop can have a non-regular behaviour, especially when it has
several paths. Some of these paths can stay impossible for a while, and become
possible after a certain number of iterations. However, the widening operator
extrapolates without taking care of these new possible paths. This can induce a
loss of precision, that Lookahead widening tries to limit.

@@

	\subsubsection{Reachability problem}
	\label{reachability}
	Path focusing technique tries to discover which paths to focus on for the
	computation of the fixpoint iterations. 
	Since the multigraph is not constructed explicitly, the different paths
	starting and finishing on node in $P_R$ are not known. 
	The technique proposes to express the focus path as the solution of a
	satisfiability modulo theory (SMT) problem:

@@

	We apply the \emph{Path Focusing} technique to our running example from
	Figures \ref{runningexample}, \ref{multigraphtransformGopan} and
	\ref{multigraphGopan}.
	The control points in $P_R$ are $p_0$, $p_1$ and $p_7$. 
	The $\rho$ formula associated to this program is depicted in
	Figure \ref{rhoformula}.
	One computes
	iteratively the associate abstract values $X_0$, $X_1$ and $X_7$.

@@

	The technique is aimed at computing a disjunctive invariant for this control
	point: 
	$$\displaystyle\bigvee_{1 \leq j \leq m} X_{p,j}$$
	where $X_{p,j}$ is a conjunction of linear inequalities, i.e a convex
	polyhedra.

@@

	\subsection{Removing Identity transitions}

	When computing iterations, some of the loop paths may not change the state
	of the variables, i.e the transition function associated to this path is the
	identity function. This happens in the case of expansive transition
	relations $\Phi$, defined by the property $X \subseteq \Phi(X)$.

	Considering the set of initial states $I$, and the transition function 
	$\tau$ mapping a state to its successor, we can define:

	$$\Phi(X) = \{ x'\ |\ \exists x \in X, x' = \tau(x)\} \cup I$$

	$\Phi$ is the function we classically use to find an inductive invariant
	$X$,satisfying the property $\Phi(X) \subseteq X$. To improve the analysis,
	we propose to compute this invariant using another function $\Psi$, by
	removing the transitions that are actually the identity. Indeed, these
	transitions make no change to the set $X$.
	The $\Psi$ function could be expressed as follows:

	$$\Psi(X) = \{ x'\ |\ \exists x \in X, x' = \tau(x) \wedge x' \neq x\}
	\cup I$$

	It is easy to prove that $\Psi(X) \subseteq X \Leftrightarrow \Phi(X)
	\subseteq(X)$, so we can use $\Psi$ instead of $\Phi$ to compute our
	invariant.

	In the case of path focusing algorithm, we simply conjunct our formula with
	$(i \neq j) \vee (x'_j \neq x_i)$, such that the SMT-solving only proposes
	paths that are not the identity.

@@

Another example of information that was missing is the set of live variables.
Indeed, in LLVM, live variables computation is processed after the register
allocation. Then, the data structure we work on doesn't have live variables
informations yet. To solve this problem, we coded our own pass, that is aimed at
computing these live variables after the transformation in SSA form.

@@

	Validation of softwares has become an essential domain in industry,
	especially those dealing with embedded systems, where a bug in the program
	can endanger the safety of people, or induce huge costs to fix. These
	domains constitute a privileged application field for formal verification
	techniques.
	Abstract Interpretation is a common used technique to do static analysis of
	programs. It is aimed at computing iteratively an over-approximation of the set of
	possible states of the program, and allows to prove properties such as
	``In this program, $x$ is always between 1 and 10'', ``The program never
	divides by zero'', etc. The objective is to obtain a result as precise as
	possible, by limiting the loss of precision during the analysis, induced for
	instance by the widening operator, or control-flow merges.
